---
title: "Intro to Rshiny"
author: "Qin Huang"
date: "2025-05-12"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(dplyr)
```

## Introduction

**Objective:** Understand what R Shiny is and what it's used for.

### What is Shiny?

R Shiny is an R package developed by Posit (formerly RStudio) that allows you to build interactive web applications directly from R --- no need to know HTML, CSS, or JavaScript.

It's a powerful tool for turning R code, data analyses, and visualizations into responsive user interfaces.

### Key features

Feature \| Description

💻 Built with R \| Create web apps entirely using R

🔄 Interactive \| Reacts instantly to user input without reloading the page

📈 Visualization \| Seamless integration with plotting libraries like ggplot2, plotly

📊 Data-driven \| Ideal for data dashboards, reports, and model interfaces

🌍 Web deployment \| Deploy via Shiny Server, shinyapps.io, or RStudio Connect

### Installing Shiny

```{r}
# Install Shiny if not already installed
# install.packages("shiny")

# Load the package

library(shiny)
```

### Basic Structure of a Shiny App

Every Shiny app has three key components:

1.  **UI (User Interface):**\

    Defines the layout and appearance of the app --- what the user sees.

2.  **Server Function:**\

    Contains the logic for handling user input, generating outputs (text, plots, tables), etc.

3.  **shinyApp():**\

    A function that puts everything together to launch the app.

Here is a small example:

```{r}

ui <- fluidPage(
  titlePanel("My First Shiny App"),
  sidebarLayout(
    sidebarPanel(
      sliderInput("num", "Choose a number:", 1, 100, 50)
    ),
    mainPanel(
      textOutput("result")
    )
  )
)

server <- function(input, output) {
  output$result <- renderText({
    paste("You selected:", input$num)
  })
}

shinyApp(ui = ui, server = server)

```

## Basic functions

### textInput

The basic structure of textInput is as below:

`textInput(inputId,label,value="")`

where `inputId` is the name server will refer to, `label` is the content showing in the ui panel, `value` is the default value.

Here's an example, we want to create a small page to present personal information the user input.

```{r}
ui <- fluidPage(
  titlePanel("Test textInput"),
  sidebarLayout(
    sidebarPanel(
      h4("Enter the personal information"),
      textInput("name", "Enter your name:"),
      textInput("age", "Enter your age:")
    ),
    mainPanel(
      h3('Personal information'),
      textOutput("nameout"),
      textOutput('ageout')
    )
  )
)

server <- function(input, output) {
  output$nameout <- renderText(input$name)
  output$ageout <- renderText(input$age)
}

shinyApp(ui = ui, server = server)
```

### numericInput

The basic structure of numericInput is as below:

`numericInput(inputId,label,value,min=NA,max=NA,step=NA)`

where `inputId` is the name server will refer to, `label` is the content showing in the ui panel, `value` is the initial value, `min` is minimum allowed value, `max` is maximum allowed value, `step` is the interval to use when stepping between min and max.

Here's an example,

```{r}
ui <- fluidPage(
  titlePanel("Numeric Input Example"),
  numericInput("num", "Choose a number:", value = 10, min = 1, max = 100),
  textOutput("numout")
)

server <- function(input, output) {
  output$numout <- renderText(paste("You chose number:", input$num))
}

shinyApp(ui = ui, server = server)
```

### radioButtons

The basic structure of radioButtons is as below:

`radioButtons(inputId,label,choices=NULL,selected=NULL)`

where `inputId` is the name server will refer to, `label` is the content showing in the ui panel, `choice` is the list of values to select from (the values should be strings), `selected` is the initially selected value.

Here's an example,

```{r}
ui <- fluidPage(
  titlePanel("Radio Buttons Example"),
  radioButtons("gender", "Select Gender:", choices = c("Male", "Female", "Prefer not to answer")),
  textOutput("genderout")
)

server <- function(input, output) {
  output$genderout <- renderText(paste("Gender:", input$gender))
}

shinyApp(ui = ui, server = server)
```

### checkboxInput

The basic structure of checkboxInput is as below:

`checkboxInput(inputId,label,value=FALSE)`

where `inputId` is the name server will refer to, `label` is the content showing in the ui panel, `value` is the initial value (true/false).

Here's an example,

```{r}
ui <- fluidPage(
  titlePanel("Checkbox Input Example"),
  checkboxInput("subscribe", "Subscribe to newsletter", value = FALSE),
  textOutput("subscribeout")
)

server <- function(input, output) {
  output$subscribeout <- renderText({
    if (input$subscribe) "Subscribed" else "Not Subscribed"
  })
}

shinyApp(ui = ui, server = server)
```

### checkboxGroupInput

The basic structure of checkboxInput is as below:

`checkboxInput(inputId,label,choices=NULL,selected=NULL)`

where `inputId` is the name server will refer to, `label` is the content showing in the ui panel, `choices` is the list of values to show checkboxes for, `selected` is the values that could be initially selected.

Here's an example,

```{r}
ui <- fluidPage(
  titlePanel("Checkbox Group Input Example"),
  checkboxGroupInput("hobbies", "Select hobbies:",
                     choices = c("Reading", "Traveling", "Gaming")),
  textOutput("hobbiesout")
)

server <- function(input, output) {
  output$hobbiesout <- renderText(paste("Hobbies:", paste(input$hobbies, collapse = ", ")))
}

shinyApp(ui = ui, server = server)
```

### sliderInput

Usage of sliderInput is as:

`sliderInput(inputId,label,min,max,value,step=NUL)`

where `inputId` is the name server will refer to, `label` is the content showing in the ui panel, `value` is the initial value, `min` is minimum allowed value, `max` is maximum allowed value, `value` is the initial value of the slider, `step` is the interval to use when stepping between min and max.

Here's an example,

```{r}
ui <- fluidPage(
  titlePanel("Slider Input Example"),
  sliderInput("range", "Choose a range:", min = 0, max = 100, value = c(20, 80)),
  textOutput("rangeout")
)

server <- function(input, output) {
  output$rangeout <- renderText(paste("Range selected:", paste(input$range, collapse = " to ")))
}

shinyApp(ui = ui, server = server)
```

### selectInput

Usage of selectInput is as:

`selectInput(inputId,label,choices,selected=NULL,multiple=FALSE)`

where `inputId` is the name server will refer to, `label` is the content showing in the ui panel, `value` is the initial value, `choices` is the list of values to select from, `selected` is the values that could be initially selected, `multiple` is the parameter to determine whether selection of multiple items is allowed.

Here's an example,

```{r}
ui <- fluidPage(
  titlePanel("Select Input Example"),
  selectInput("country", "Choose a country:",
              choices = c("USA", "Canada", "UK")),
  textOutput("countryout")
)

server <- function(input, output) {
  output$countryout <- renderText(paste("Selected country:", input$country))
}

shinyApp(ui = ui, server = server)
```

### fileInput

Usage of fileInput is as:

`fileInput(inputId,label,multiple=FALSE,accept=NULL,buttonLabel="Browse..")`

where `inputId` is the name server will refer to, `label` is the content showing in the ui panel, `multiple` is the parameter to determine whether the user should be allowed to select and upload multiple files at once, `accept` is used to restrict the type of upload file(s), `buttonLabel` is the label used on the button.

Here's an example,

```{r}
ui <- fluidPage(
  titlePanel("File Input Example"),
  fileInput("file", "Upload a file"),
  textOutput("fileout")
)

server <- function(input, output) {
  output$fileout <- renderText({
    if (is.null(input$file)) {
      "No file uploaded"
    } else {
      paste("Uploaded file:", input$file$name)
    }
  })
}

shinyApp(ui = ui, server = server)
```

### plotOutput

**Usage:**

`plotOutput(outputId, width = "100%", height = "400px", click = NULL, dblclick = NULL, hover = NULL, brush = NULL)`

**Description:**

`plotOutput` is used in the **UI** to create a placeholder for a plot generated on the **server side** using `renderPlot()`. This function is useful for displaying static or interactive plots generated using R's base plotting system or `ggplot2`.

Parameters:

-   `outputId`: A string that identifies the output and links it to the corresponding `renderPlot()` in the server.

-   `width`: Width of the plot output (can be `"100%"`, `"400px"`, etc.).

-   `height`: Height of the plot (default is `"400px"`).

-   `click`, `dblclick`, `hover`, `brush`: Optional input event handlers that allow interaction with the plot (e.g., detecting where the user clicked).

**Example**

```{r}
ui <- fluidPage(
  titlePanel("plotOutput Example"),
  plotOutput("myplot")
)

server <- function(input, output) {
  output$myplot <- renderPlot({
    plot(rnorm(100), rnorm(100), col = "blue", pch = 19,
         main = "Random Scatter Plot")
  })
}

shinyApp(ui = ui, server = server)
```

### tableOutput

**Usage**:

`tableOutput(outputId)`

**Description:**

`tableOutput` is used in the **UI** to display data frames or matrices as HTML tables. On the **server side**, it must be linked with `renderTable()` to define the content of the table. It\'s ideal for displaying clean and simple tabular data in a reactive application.

**Parameters:**

`outputId`: A string identifying the table output slot and linking it with `renderTable()` on the server.

> Note: For more interactive tables (e.g., with sorting or filtering), use `DT::dataTableOutput()` instead.

**Example:**

```{r}
ui <- fluidPage(
  titlePanel("tableOutput Example"),
  tableOutput("mytable")
)

server <- function(input, output) {
  output$mytable <- renderTable({
    head(mtcars, 5)
  })
}

shinyApp(ui = ui, server = server)
```

### uiOutput

**Usage**:

`uiOutput(outputId)`

**Description:**

`uiOutput` is used in the **UI** to create a dynamic UI element that is generated on the **server side** using `renderUI()`. This is helpful when your UI needs to change based on user input --- for example, displaying different types of inputs, conditional panels, or layout components depending on previous selections.

**Parameters:**

`outputId`: A string that identifies the output and links it to the corresponding `renderUI()` function in the server.

**Example:**

```{r}
ui <- fluidPage(
  titlePanel("uiOutput Example"),
  radioButtons("inputType", "Choose input type:",
               choices = c("Text", "Number")),
  uiOutput("dynamicInput"),
  textOutput("valueout")
)

server <- function(input, output) {
  output$dynamicInput <- renderUI({
    if (input$inputType == "Text") {
      textInput("myinput", "Enter text:")
    } else {
      numericInput("myinput", "Enter number:", value = 0)
    }
  })

  output$valueout <- renderText({
    paste("You entered:", input$myinput)
  })
}

shinyApp(ui = ui, server = server)
```

## Reactive Expressions

### reactive()

**Usage:**

<div>

reactive({

\# some expression

})

</div>

**Description:**

-   `reactive()` is used to define **reactive expressions**, which are like formulas in Excel.

-   It **returns a reactive object** that can be used in other parts of the app (e.g., in `renderText()`, `renderPlot()`).

-   It **recalculates automatically** when any of the inputs inside it change, but only when **it's actually called**.

**Example:**

```{r}
ui <- fluidPage(
  numericInput("x", "Enter a number", 1),
  numericInput("y", "Enter another number", 1),
  textOutput("sum")
)

server <- function(input, output) {
  total <- reactive({
    input$x + input$y
  })

  output$sum <- renderText({
    paste("The sum is:", total())
  })
}

shinyApp(ui = ui, server = server)
```

### observe()

**Usage:**

<div>

observe({

\# reactive expression without output

})

</div>

**Description:**

-   `observe()` is used for **side effects** --- code that doesn\'t return a value to UI but **does something**, like logging, updating inputs, or printing.

-   It automatically tracks the reactive inputs inside it.

**Example:**

```{r}
ui <- fluidPage(
  titlePanel("observe() Example"),
  
  sidebarLayout(
    sidebarPanel(
      numericInput("num", "Enter a number:", value = 1),
      textInput("log", "System log (auto updated):", value = "")
    ),
    
    mainPanel(
      textOutput("square")
    )
  )
)

server <- function(input, output, session) {
  
  # This observe() does two things:
  # 1. Prints to the R console
  # 2. Updates a text input with a message
  observe({
    val <- input$num
    log_msg <- paste("User entered:", val, "at", Sys.time())
    
    print(log_msg)
    updateTextInput(session, "log", value = log_msg)
  })
  
  output$square <- renderText({
    paste("Square of the number is:", input$num^2)
  })
}

shinyApp(ui = ui, server = server)
```

### observeEvent()

**Usage:**

`observeEvent(eventExpr, handlerExpr, ignoreInit = FALSE)`

**Description:**

-   `observeEvent()` is a **trigger-based** version of `observe()` --- it only runs **when a specific input changes**.

-   It\'s useful for **buttons** and **event-driven logic**.

**Example:**

```{r}
ui <- fluidPage(
  numericInput("x", "Enter a number", 1),
  actionButton("go", "Print number")
)

server <- function(input, output, session) {
  observeEvent(input$go, {
    print(paste("Button clicked. x =", input$x))
  })
}

shinyApp(ui = ui, server = server)
```

### actionButton

**Usage:**

`actionButton(inputId, label, icon = NULL)`

**Description:**

-   Creates a button the user can click to trigger actions.

-   Its value is an integer that increases by 1 each time it\'s clicked.

-   Often used with `observeEvent()` or `eventReactive()`.

### conditionPanel

**Usage:**

`conditionalPanel(condition, ui_element)`

**Description:**

-   Shows or hides UI elements **based on JavaScript expressions** (not R expressions!).

-   It is used to make the interface **dynamic without reloading the page**.

-   The `condition` should be a **JavaScript condition**, e.g., `input.showExtra == true`.

**Example:**

```{r}
ui <- fluidPage(
  checkboxInput("show", "Show more options", FALSE),
  conditionalPanel(
    condition = "input.show == true",
    textInput("extra", "Extra input:")
  )
)

server <- function(input, output) {}

shinyApp(ui = ui, server = server)

```

## Resource

For information, you can check via <https://shiny.posit.co/r/getstarted/shiny-basics/lesson1/> to learn more.

### 
